---
title: "STAT 628 airline project: handling dates & times"
subtitle: "Spring 2026"
format:
  pdf:
    geometry: margin=1in
    # fontsize: 11pt
---

```{r}
#| echo: false
#| warning: false
library(tidyverse)
```

```{r}
#| output: false
wd_data = "~/Library/CloudStorage/Box-Box/stat628_airplanes/"
raw_data <- read_csv(str_c(wd_data, "airlines_2025_9.csv"))
head(raw_data)
```

# Re-code Day of the Week

DayOfWeek in airlines data has 1=Monday, 2=Tuesday, ...
However, tidyverse uses 1=Sunday, 2=Monday, ...

```{r}
###Understanding `wday`
#----Tidyverse has 1=Sunday by default
wday(1, label=TRUE, abbr = TRUE) # 1=Sunday, not Monday but Monday is first level
wday(1, label=TRUE, week_start=1) # 1 is still Sunday, 
#but documentation says it should be Monday 
#(run "?wday" and read about `week_start`)
wday(1, label=TRUE, week_start=7) #This makes Sunday the first level
wday(1:7 + 1, label=TRUE) #Need to adjust the start of the week or it makes Sunday NA
options("lubridate.week.start" = 1)
wday(1:7, label=TRUE) # 1=Sunday, but Monday < ...
wday(1:7 + 1, label=TRUE, week_start=1) #Need to +1 to match the airlines data

raw_data %>% 
  select(DayOfWeek, FlightDate) %>%
  slice(1) %>%
  mutate(wday_num = wday(DayOfWeek, label=TRUE),  #2025-09-06 was on a Saturday
         wday_date = wday(FlightDate, abbr=TRUE))

raw_data %>% 
  filter(is.na(DayOfWeek)) %>%
  nrow() # empty

raw_data %>%
  mutate(DayOfWeek = wday(DayOfWeek+1, label=TRUE)) %>% #Need to add 1 to DayOfWeek 
#to get the day correct
  group_by(DayOfWeek) %>%
  count()
```

Recode day of week, recode cancellation codes, and save in `f`light `dat`a
(don't call your data `data`, because `data` is already a function):



Here is how cancellation code is defined:
A	Carrier
B	Weather
C	National Air System (NAS)
D	Security

```{r}
fdat <- raw_data %>%
  mutate(DayOfWeek = wday(DayOfWeek+1, label=TRUE)) %>%
  mutate(CancellationCode = recode_values(CancellationCode,
    NA ~ NA,
    "A" ~ "Carrier",
    "B" ~ "Weather",
    "C" ~ "NAS",
    "D" ~ "Security"))


fdat %>%
  group_by(CancellationCode) %>%
  count()
```

Next, it can be useful to rename variables to use important ones more easily:

```{r}
fdat <- fdat %>%
  rename(
    year = "Year",
    month = "Month",
    week = "DayOfWeek",
    day = "DayofMonth",
    date = "FlightDate",
    origin_iata = "Origin", #origin airport
    origin_city = "OriginCityName",
    dest_iata   = "Dest",
    dest_city   = "DestCityName",
    airline = "Reporting_Airline", #Unique Carrier Code. 
#When the same code has been used by multiple carriers, 
#a numeric suffix is used for earlier users, for example, PA, PA(1), PA(2). 
#Use this field for analysis across a range of years.
    fl_num  = "Flight_Number_Reporting_Airline", #Flight Number
    sch_dep_time = "CRSDepTime", #CRS Departure Time (local time: hhmm)
    act_dep_time = "DepTime", #Actual Departure Time (local time: hhmm)
    sch_arr_time = "CRSArrTime", #CRS Arrival Time (local time: hhmm)
    act_arr_time = "ArrTime",  #Actual Arrival Time (local time: hhmm)
    sch_duration = "CRSElapsedTime", #CRS Elapsed Time of Flight, in Minutes
    act_duration = "ActualElapsedTime" #Elapsed Time of Flight, in Minutes
  )
```

Next, we can remove some variables not used for this date-time demo (though you may want to use them for your analysis):

```{r}
fdat <- fdat %>%
  select(!c(DOT_ID_Reporting_Airline, 
            OriginAirportID:OriginCityMarketID,
            AirTime,Diverted,Distance:TotalAddGTime))
```


# Add airport elevation and timezone

To make all dates & times comparable, we need to get the time zone of each airport, because all times are currently in the local time zone.

I show here how to add, for each airport (origin & destination: two airports per row), multiple pieces of information that you should use (and see if they do / do not help make predictions):

1. the airport time zone: in `airports_timezone.csv`
2. the airport type (small, medium, large)
3. and its elevation: in `airports_elevation.csv`

Both files and the delay data files include the IATA airport code,
which can be used as a key to join the two datasets.

IANA: Internet Assigned Numbers Authority, for time zones
IATA: International Air Transport Association, for airport codes

```{r}
#| output: false
airportinfo <- read_csv(str_c(wd_data, "airports_elevation.csv"))  

airportinfo |>  group_by(type) |>  count() # some heliport and seaplane_base

airport_tz <- 
  read_csv(str_c(wd_data,  "airports_timezone.csv")) |>
  select(iata_code, iana_tz)
```

## Add origin airport info

```{r}
airportinfo_origin <- airportinfo |>
  rename(origin_type = "type",
         origin_elev = "elevation_ft",
         origin_iata = "iata_code")

tz_origin <- airport_tz |>
  rename(origin_iata = "iata_code",
         origin_tz = "iana_tz")

fdat <- fdat |>
  left_join(airportinfo_origin, by = "origin_iata") |>
  left_join(tz_origin, by = "origin_iata")

fdat |> group_by(origin_tz) |> count() |> print(n=Inf) # looks good 
```

## Add destination airport info

```{r}
airportinfo_dest <- airportinfo |>
  rename(dest_type = "type",
         dest_elev = "elevation_ft",
         dest_iata = "iata_code")

tz_dest <- airport_tz |>
  rename(dest_iata = "iata_code",
         dest_tz = "iana_tz")

fdat <- fdat |>
  left_join(airportinfo_dest, by = "dest_iata") |>
  left_join(tz_dest, by = "dest_iata")
fdat |> group_by(dest_tz) |> count() |> print(n=21)
```

# Convert times to UTC

Dates & times are tricky (time zones, leap days, daylight savings times,
different input format). A date, a date-time, and a duration are different things.

Here we have to deal with different time zones and red-eye flights.
Once handled correctly, we can subtract time-dates, or add a time-date and a duration.

Use [lubridate](https://lubridate.tidyverse.org) for this!

What we have:

```{r}
fdat |>
  select(year,month,day,date,sch_dep_time,origin_tz, sch_arr_time,dest_tz) |>
  head()
```

A simple example:

```{r}
?make_datetime
tmp = make_datetime(year=2025, 
                    month=5, 
                    day=1,
                    hour=as.integer(substr("1615",1,2)),
                    min=as.integer(substr("1615",3,4)),
                    sec=0,
                    tz = "America/New_York") 
#"America/New_York", "America/Chicago", "America/Los_Angeles"

tmp

with_tz(tmp, tzone="America/Chicago") #with_tz returns a date-time as it would 
#appear in a different time zone.
with_tz(tmp, tzone="UTC") 
```


Alternatively, no `tz` argument in `make_datetime` then give it later to `force_tz`:
```{r}
tmp = make_datetime(year=2025, 
                    month=5, 
                    day=1,
                    hour=as.integer(substr("1615",1,2)),
                    min=as.integer(substr("1615",3,4)),
                    sec=0) # no tz argument
tmp

force_tz(tmp, tzone="America/New_York")

with_tz(force_tz(tmp, tzone="America/New_York"), "UTC")
```


We will do the above on our data, but we will need the hour and minutes
separately, and as numeric variables, not characters.

```{r}
fdat <- fdat |> mutate(
  sch_dep_hour = as.integer(substr(sch_dep_time,1,2)),
  sch_dep_min  = as.integer(substr(sch_dep_time,3,4)),
  act_dep_hour = as.integer(substr(act_dep_time,1,2)),
  act_dep_min  = as.integer(substr(act_dep_time,3,4)),
  sch_arr_hour = as.integer(substr(sch_arr_time,1,2)),
  sch_arr_min  = as.integer(substr(sch_arr_time,3,4)),
  act_arr_hour = as.integer(substr(act_arr_time,1,2)),
  act_arr_min  = as.integer(substr(act_arr_time,3,4))
)
```

Sanity checks (always do!):
```{r}
fdat |> count(sch_dep_hour) |> print(n=25)

fdat |> filter(is.na(sch_dep_time)) |> nrow() # empty: great

str_pad(as.character(c(7,17,117)), width=2, pad="0") # "07"  "17"  "117"

fdat |> select(sch_dep_time, sch_dep_hour, sch_dep_min) |>
  mutate(
    tmp = paste0(str_pad(as.character(sch_dep_hour),width=2,pad="0"), #Make departure 
                 #hours and minutes width=2
                 str_pad(as.character(sch_dep_min), width=2,pad="0")),
    bad = is.na(sch_dep_hour) | is.na(sch_dep_min) | (sch_dep_time != tmp) ) |> 
  #Make sure tmp == sch_dep_time
  filter(bad) |> nrow() # empty: great!
```

**Caveat**: red-eye flights have a scheduled arrival day *different* from `day`.
To get their scheduled arrival date-time (in UTC), we should use their
arrival time from the departure + duration, instead of what we do above & below.
We will fix that later.

Immediate **caveat**: `make_datetime` does *not* use its `tz` argument
correctly within `mutate`, without any warning. It seems to be using the
timezone of the first row only, not the row's time zone from that row.
But `force_tz` seems to be using its `tzone` argument correctly within `mutate`:

Note: UTC is 4h ahead of NY, 5h ahead of Chicago, 6h ahead of Denver, 7h ahead of LA.
Note 2: "Phoenix, Arizona, is consistently 7 hours behind Coordinated Universal Time (UTC-7) year-round. Because Arizona does not observe Daylight Saving Time (DST), this 7-hour difference remains constant, never switching to UTC-6, unlike other Mountain Time locations."
```{r}
fdat |>
  mutate(
    sch_dep_time_utc1 = with_tz(make_datetime( ###########make_datetime (fails!)
      year, month, day, sch_dep_hour, sch_dep_min, 0, origin_tz),
      tzone="UTC"),
    sch_dep_time_utc2 = with_tz(force_tz(make_datetime( ###########force_tz (works!)
      year, month, day, sch_dep_hour, sch_dep_min, 0), #no "tzone" specified
      tzone=origin_tz),tzone="UTC")
  ) |>
  select(day, sch_dep_time,origin_tz,
    sch_dep_time_utc1,sch_dep_time_utc2) |>
  # head() #First row = America/Chicago
  filter(sch_dep_time_utc1 != sch_dep_time_utc2) |> 
  print(n=5)
```


We will use the second version that has no `tz` argument in `make_datetime`,
but applies `force_tz` with the origin / destination airport time zone.

Lesson: make **lots of sanity checks** to avoid bugs, sometimes big ones.

```{r}
fdat <- fdat |>
  mutate(
    sch_dep_time_utc = with_tz(force_tz(make_datetime(
      year, month, day, sch_dep_hour, sch_dep_min, 0),
      tzone=origin_tz),tzone="UTC"),
    act_dep_time_utc = with_tz(force_tz(make_datetime(
      year, month, day, act_dep_hour, act_dep_min, 0),
      tzone=origin_tz),tzone="UTC"),
    dep_delay_time_utc = act_dep_time_utc - sch_dep_time_utc, 
    # drtn (Duration): Measured in seconds
  )

fdat %>% 
  select(sch_dep_time,act_dep_time, dep_delay_time_utc) %>% 
  head()
```

Very basic sanity check:
```{r}
fdat |>
  mutate(
    dep_delay_sec = DepDelay * 60, ###DepDelay is in minutes
    dep_delay_min = as.integer(dep_delay_time_utc)/60) |> ###dep_delay_time_utc is a duration in seconds
  select(DepDelay, dep_delay_sec, dep_delay_time_utc, dep_delay_min) |>
  head()
```



A more thorough check, including how NAs were handled.
```{r}
tmp <- fdat |>
  mutate(
    dep_delay_min = as.integer(dep_delay_time_utc)/60, #seconds to minutes
    dep_delay_diff = DepDelay - dep_delay_min) #check diff in our calc vs data
summary(tmp$DepDelay) #Recall:  in minutes
summary(tmp$dep_delay_min)  # fewer NAs...
summary(tmp$dep_delay_diff) # not all 0s...

#Below we look at rows with NA DepDelay, but our calc not NA - all were canceled
tmp |>
  filter(is.na(DepDelay) & !is.na(dep_delay_min))  |>  #View() 
  #We computed a delay, when DepDelay was NA
  group_by(Cancelled) |>
  #count()
  summarize(dep_delay_min = sum(dep_delay_min)) # all cancelled

#Next check differences in (non-NA) raw vs calculated delays, when delays present 
tmp |> filter(!is.na(DepDelay) & (dep_delay_diff != 0)) |>
  select(sch_dep_time, act_dep_time, DepDelay, dep_delay_diff) |>
  # View()
  group_by(dep_delay_diff) |> count() # -1, 1 or 2 times 1440min = 24 hours


tmp |> 
  filter(Cancelled == 1 & !is.na(DepDelay)) |> 
  group_by(CancellationCode) |> 
  count() #Flights were canceled after they left the gate 
```

Conclusion:

- Do not use our own calculated delay, since some flights were canceled after being delayed

  

```{r}
fdat <- fdat |> mutate(dep_delay_time_utc=NULL) #This removes the noted column
```



## Arrival times for red-eye flights

Scheduled arrival time should be: scheduled departure time + scheduled duration.

```{r}
summary(fdat$sch_duration) # numeric already, in minutes
ymd_hms("2025-05-01 16:15:00 EDT")
ymd_hms("2025-05-01 16:15:00 EDT") + 10 # 10 interpreted in seconds
ymd_hms("2025-05-01 16:15:00 EDT") + 10*60
```

Repeat from above, but with arrival instead of departure:

```{r}
fdat <- fdat |>
  mutate(
    sch_arr_time_utc_tmp = with_tz(force_tz(make_datetime(
      year, month, day, sch_arr_hour, sch_arr_min, 0),
      tzone=dest_tz),tzone="UTC"),
    sch_arr_time_utc = sch_dep_time_utc + (sch_duration * 60) 
  )
```

Check:
```{r}
fdat |> select(sch_arr_time_utc, sch_arr_time_utc_tmp) |>
  mutate(diff = sch_arr_time_utc_tmp - sch_arr_time_utc) |>
  group_by(diff) |> count()  # most 0, many -86400s=-1 day, a few other values
```

These may be actual errors in the input data:
```{r}
fdat |> mutate(diff = sch_arr_time_utc_tmp - sch_arr_time_utc) |>
  filter(diff != 0 & diff !=-86400 & diff !=86400) |>
  select(day,origin_city,origin_tz,dest_city,dest_tz,
         DepDelay, TaxiOut, WheelsOff, TaxiIn, WheelsOn, sch_duration,
    sch_dep_time,sch_dep_hour,sch_dep_min,
    sch_arr_time,sch_duration,sch_dep_time_utc,
    sch_arr_time_utc_tmp,sch_arr_time_utc, diff
  ) |>
  slice(3) |> 
  print(width=200) |> 
  t()
```


One example:
Flight from Atlanta to Dallas has a scheduled duration of 86 minutes.  It seems this flight should actually take 125 - 145 minutes (fastest recorded was 115 minutes).  Therefore, it seems the `sch_arr_time_utc` is off.
Furthermore, if the `sch_duration` of 86 minutes was correct, the `sch_arr_time` of "0726" is not in the expected local time, but is still in the departure time zone.

day                  "23"                   
origin_city          "Atlanta, GA"          
origin_tz            "America/New_York"     
dest_city            "Dallas/Fort Worth, TX"
dest_tz              "America/Chicago"      
DepDelay             "-9"                   
TaxiOut              "9"                    
WheelsOff            "0600"                 
TaxiIn               "11"                   
WheelsOn             "1042"                 
sch_duration         "86"                   
sch_dep_time         "0600"                 
sch_dep_hour         "6"                    
sch_dep_min          "0"                    
sch_arr_time         "0726"                 
sch_dep_time_utc     "2025-09-23 10:00:00"  
sch_arr_time_utc_tmp "2025-09-23 12:26:00"  
sch_arr_time_utc     "2025-09-23 11:26:00"  
diff                 "3600 secs"  

You can see 86 is an unusual duration for this flight:
```{r}
fdat |> mutate(diff = sch_arr_time_utc_tmp - sch_arr_time_utc) |>
  filter(origin_city=="Atlanta, GA", dest_city=="Dallas/Fort Worth, TX") |>
  select(day,origin_city,origin_tz,dest_city,dest_tz,
         sch_duration) %>%
  group_by(sch_duration) %>% 
  count()
```



